---
title: "ATACseq analysis with Itokawa Aging data"
author: "Julia Hauenstein"
date: "`r format(Sys.Date(),format='%d/%m/%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

## Load packages needed
To install either use: install.packages("package1") or BiocManager::install("package2")

```{r install_packages, message=FALSE, error=FALSE}
library(tibble) # Needed to work with tibbles
library(dplyr) # Needed to work with tibbles
library(magrittr) # Needed for pipes
library(preprocessCore) # Needed for logquant normalisation
library(ggplot2) # Needed for plotting
library(ggrepel) # Needed if labels are added to the plots
library(pheatmap) # Needed for plotting heatmaps
library(limma) # Needed to correct for batch effect
library(DESeq2) # Needed for differential analysis
library(VennDiagram) # Needed to make Venn diagrams
```

## Load file with defined functions

```{r load_functions}
source("HSC_Aging_analysis.fun.R")
```

## Set colours for plotting

```{r set_colours}
colours_Celltypes <- c("#DCBF30","#9E1F19","#ffbead","#300740","#CA33FF","#2F62DA","#33E6FF")
colours_Age <- c("#EC7B22", "#6DB7DA","#074010")
```

## Load data

```{r load_input_5Adult}
df_raw <- read.delim("../data/ATACseq_analysis/mm10_230910_ATAC_Aging_Itokawa_2020_macspeaks_tagdir_mergedPeaks_noadj.txt", header=TRUE, sep="\t")
basic <- df_raw[,1:19]

annotation <- read.csv2("../data/ATACseq_analysis/mm10_230910_ATAC_Aging_Itokawa_2020_macspeaks_tagdir_annotations.csv", header=TRUE) %>% as_tibble()
annotation <- annotation %>%
  mutate(Cell_type = factor(Cell_type, levels=c("CD49bneg", "CD49bpos", "HSC_I", "LMPP", "LMPP_I", "GMP", "GMP_I"))) %>%
  mutate(AgeCelltype = paste(Age, Cell_type, sep="_")) %>%
  mutate(Age = factor(Age, levels=c("Juvenile","Adult","Old")))

# Define indexes for CD150high samples
i_main <- which(annotation$Cell_type=="CD49bneg" | annotation$Cell_type=="CD49bpos" | annotation$Cell_type=="HSC_I")
i_main_df <- c(1:length(basic), length(basic) + i_main)
```

## Peak filtering based on RPKM
More than 5RPKM in at least 1/3 of a population are required for a peak to be considered found

```{r RPKM_filter}
df_RPKM <- RPKMs(df_raw, annotation, basic)
 
peaks_found_in_CD49bpos_Old <- peaks_found(df_RPKM, annotation, "Old_CD49bpos")
peaks_found_in_CD49bneg_Old <- peaks_found(df_RPKM, annotation, "Old_CD49bneg")
peaks_found_in_LMPP_Old <- peaks_found(df_RPKM, annotation, "Old_LMPP")
peaks_found_in_GMP_Old <- peaks_found(df_RPKM, annotation, "Old_GMP")
peaks_found_in_CD49bpos_Adult <- peaks_found(df_RPKM, annotation, "Adult_CD49bpos")
peaks_found_in_CD49bneg_Adult <- peaks_found(df_RPKM, annotation, "Adult_CD49bneg")
peaks_found_in_LMPP_Adult <- peaks_found(df_RPKM, annotation, "Adult_LMPP")
peaks_found_in_GMP_Adult <- peaks_found(df_RPKM, annotation, "Adult_GMP")
peaks_found_in_CD49bpos_Juvenile <- peaks_found(df_RPKM, annotation, "Juvenile_CD49bpos")
peaks_found_in_CD49bneg_Juvenile <- peaks_found(df_RPKM, annotation, "Juvenile_CD49bneg")
peaks_found_in_LMPP_Juvenile <- peaks_found(df_RPKM, annotation, "Juvenile_LMPP")
peaks_found_in_GMP_Juvenile <- peaks_found(df_RPKM, annotation, "Juvenile_GMP")

peaks_found_in_HSC_Adult_I <- peaks_found(df_RPKM, annotation, "Adult_HSC_I")
peaks_found_in_HSC_Old_I <- peaks_found(df_RPKM, annotation, "Old_HSC_I")
peaks_found_in_LMPP_Adult_I <- peaks_found(df_RPKM, annotation, "Adult_LMPP_I")
peaks_found_in_LMPP_Old_I <- peaks_found(df_RPKM, annotation, "Old_LMPP_I")
peaks_found_in_GMP_Adult_I <- peaks_found(df_RPKM, annotation, "Adult_GMP_I")
peaks_found_in_GMP_Old_I <- peaks_found(df_RPKM, annotation, "Old_GMP_I")

peaks_kept_merged <- union(union(union(union(union(union(union(union(union(union(union(union(union(union(union(union(union(peaks_found_in_CD49bpos_Old,peaks_found_in_CD49bneg_Old),peaks_found_in_LMPP_Old),peaks_found_in_GMP_Old),peaks_found_in_CD49bpos_Adult),peaks_found_in_CD49bneg_Adult),peaks_found_in_LMPP_Adult),peaks_found_in_GMP_Adult),peaks_found_in_CD49bpos_Juvenile),peaks_found_in_CD49bneg_Juvenile),peaks_found_in_LMPP_Juvenile),peaks_found_in_GMP_Juvenile),peaks_found_in_HSC_Adult_I),peaks_found_in_HSC_Old_I),peaks_found_in_LMPP_Adult_I),peaks_found_in_LMPP_Old_I),peaks_found_in_GMP_Adult_I),peaks_found_in_GMP_Old_I)

df_raw_filtered <- df_raw[peaks_kept_merged,]
```

## Normalize using log quantile transformation

```{r LogQuant_normalize}
df_LQT <- LogQuantTrans(df_raw_filtered, annotation, basic)
```

## Batch correction
```{r}
### Batch correction ###
batch_correction <- function(df, annotation, basic){
  my_batch <- as.factor(annotation$batch)

  countsCols <- c((length(basic)+1):(length(c(basic,rownames(annotation)))))
  data <- data.matrix(df[,countsCols])
  data_batchcorrected <- removeBatchEffect(data, batch=my_batch)
  df[,countsCols] <- data_batchcorrected
  return(df)
}

annotation$batch <- "Luc"
annotation$batch[which(annotation$Flow_cell == "Itokawa")] <- "Itokawa"
df_LQT_corrected <- batch_correction(df_LQT, annotation, basic)
```

## QC plots

```{r QC_plots}
# Boxplot
boxplt(df_LQT, annotation, basic, "All samples")
boxplt(df_LQT_corrected, annotation, basic, "All samples - corrected")
```

## PCA plots
Supplementary Figure S8b

```{r PCA_plots}
p <- PCAplt(df_LQT_corrected, annotation, basic, groupVector=annotation$Cell_type, groupVector_shape=annotation$Age, plotName="All samples - PC1 vs. PC2", showFilename=FALSE, pc1=1, pc2=2, colorVector=colours_Celltypes)

setEPS()
postscript("../plots/ATACseq_analysis/S8b_PCA_allPopulations_with_Itokawa.eps", width=6, height=5)
  PCAplt(df_LQT_corrected, annotation, basic, groupVector=annotation$Cell_type, groupVector_shape=annotation$Age, plotName="All samples - PC1 vs. PC2", showFilename=FALSE, pc1=1, pc2=2, colorVector=colours_Celltypes)
dev.off()
```

## Correlation heatmap
Supplementary Figure S8c

```{r Correlation_heatmap, message=FALSE}
Correlation_Heatmap(df_LQT_corrected, annotation, basic, "All cell populations", cor="spearman", colours_Celltypes=colours_Celltypes, colours_Age=colours_Age)

setEPS()
postscript("../plots/ATACseq_analysis/S8c_CorrelationHeatmap_allPopulations_with_Itokawa.eps", width=8, height=7)
  Correlation_Heatmap(df_LQT_corrected, annotation, basic, "All cell populations", cor="spearman", colours_Celltypes=colours_Celltypes, colours_Age=colours_Age)
dev.off()
```

# Load reagions that are diff in Juvenile vs. Old in our data or that are diff in CD49b+ vs CD49bneg
```{r}
JvsO_cluster1 <- read.delim("../output_files/ATACseq_analysis/Diff_merged_J_O_main_clust1_of_3.bed", header=FALSE, sep="\t")

gr_JvsO_cluster1 <- GRanges(seqnames = as.character(JvsO_cluster1$V1),
                            ranges = IRanges(start=JvsO_cluster1$V2,
                                             end=JvsO_cluster1$V3,
                                             names=JvsO_cluster1$V4))

gr_df_raw_filtered <- GRanges(seqnames = as.character(df_raw_filtered$Chr),
                            ranges = IRanges(start=df_raw_filtered$Start,
                                             end=df_raw_filtered$End,
                                             names=df_raw_filtered$Location))

overlaps_JvsO_cluster1 <- findOverlaps(query=gr_JvsO_cluster1, subject=gr_df_raw_filtered)
df_LQT_corrected_cluster1 <- df_LQT_corrected[subjectHits(overlaps_JvsO_cluster1),]

YvsO_cluster2 <- read.delim("../output_files/ATACseq_analysis/Diff_merged_J_O_main_clust2_of_3.bed", header=FALSE, sep="\t")

gr_YvsO_cluster2 <- GRanges(seqnames = as.character(YvsO_cluster2$V1),
                            ranges = IRanges(start=YvsO_cluster2$V2,
                                             end=YvsO_cluster2$V3,
                                             names=YvsO_cluster2$V4))

overlaps_YvsO_cluster2 <- findOverlaps(query=gr_YvsO_cluster2, subject=gr_df_raw_filtered)
df_LQT_corrected_cluster2 <- df_LQT_corrected[subjectHits(overlaps_YvsO_cluster2),]

YvsO_cluster3 <- read.delim("../output_files/ATACseq_analysis/Diff_merged_J_O_main_clust3_of_3.bed", header=FALSE, sep="\t")

gr_YvsO_cluster3 <- GRanges(seqnames = as.character(YvsO_cluster3$V1),
                            ranges = IRanges(start=YvsO_cluster3$V2,
                                             end=YvsO_cluster3$V3,
                                             names=YvsO_cluster3$V4))

overlaps_YvsO_cluster3 <- findOverlaps(query=gr_YvsO_cluster3, subject=gr_df_raw_filtered)
df_LQT_corrected_cluster3 <- df_LQT_corrected[subjectHits(overlaps_YvsO_cluster3),]
```

# Heatmap
Supplementary Figure S8e

```{r}
df_LQT_corrected_all_clusters <- rbind(df_LQT_corrected_cluster1,df_LQT_corrected_cluster2,df_LQT_corrected_cluster3)

df_LQT_corrected_all_clusters[,i_main_df] %>%
  Row_norm(annotation[i_main,], basic) %>%
  Heat_map(annotation[i_main,], basic, "Aging comparison including Itokawa",
           rownames = FALSE, ycluster = FALSE, xcluster = FALSE, colours_Celltypes = colours_Celltypes[1:3], colours_Age = colours_Age, order = "Age")

setEPS()
postscript("../plots/ATACseq_analysis/S8e_Heatmap_Diff_JuvenilevsOld_with_Itokawa.eps", width=9, height=6)
df_LQT_corrected_all_clusters[,i_main_df] %>%
  Row_norm(annotation[i_main,], basic) %>%
  Heat_map(annotation[i_main,], basic, "Aging comparison including Itokawa",
           rownames = FALSE, ycluster = FALSE, xcluster = FALSE, colours_Celltypes = colours_Celltypes[1:3], colours_Age = colours_Age, order = "Age")
dev.off()
```

# Regions diff between CD49bneg and CD49bpos - Heatmap
Supplementary Figure S9b

```{r}
annotation$Cell_type <- factor(annotation$Cell_type, levels = c("CD49bneg", "HSC_I", "CD49bpos", "LMPP", "LMPP_I", "GMP", "GMP_I"))

Up_Oneg_Opos <- read.delim("../output_files/ATACseq_analysis/Up_Oneg_Opos.bed", header=FALSE, sep="\t")

gr_Up_Oneg_Opos <- GRanges(seqnames = as.character(Up_Oneg_Opos$V1),
                            ranges = IRanges(start=Up_Oneg_Opos$V2,
                                             end=Up_Oneg_Opos$V3,
                                             names=Up_Oneg_Opos$V4))

overlaps_Up_Oneg_Opos <- findOverlaps(query=gr_Up_Oneg_Opos, subject=gr_df_raw_filtered)
df_LQT_corrected_Up_Oneg_Opos <- df_LQT_corrected[subjectHits(overlaps_Up_Oneg_Opos),]

Down_Oneg_Opos <- read.delim("../output_files/ATACseq_analysis/Down_Oneg_Opos.bed", header=FALSE, sep="\t")

gr_Down_Oneg_Opos <- GRanges(seqnames = as.character(Down_Oneg_Opos$V1),
                            ranges = IRanges(start=Down_Oneg_Opos$V2,
                                             end=Down_Oneg_Opos$V3,
                                             names=Down_Oneg_Opos$V4))

overlaps_Down_Oneg_Opos <- findOverlaps(query=gr_Down_Oneg_Opos, subject=gr_df_raw_filtered)
df_LQT_corrected_Down_Oneg_Opos <- df_LQT_corrected[subjectHits(overlaps_Down_Oneg_Opos),]

# Final heatmap only regions diff in Old
Heat_map_3 <- function(df, annotation, basic, plotName, colours_Celltypes, colours_Age) {
  countsCols <- c((length(basic)+1):(length(c(basic,rownames(annotation)))))
  
  myorder <- order(annotation$Age, annotation$Cell_type)
  group_annotation <- data.frame(Population = as.character(annotation$Cell_type[myorder]),
                                 Age = as.character(annotation$Age[myorder]))
  row.names(group_annotation) <- colnames(df[,countsCols[myorder]])
  names(colours_Celltypes) <- unique(annotation$Cell_type[myorder])
  names(colours_Age) <- unique(annotation$Age[myorder])
  my_colours <- list(Population = colours_Celltypes,
                     Age = colours_Age)
  
  pheatmap(df[,countsCols[myorder]], cluster_cols = FALSE, cluster_rows = FALSE,
           main = plotName, show_colnames = FALSE, annotation_col = group_annotation,
           show_rownames = FALSE, fontsize_row = 2, annotation_colors = my_colours,
           clustering_distance_rows = "euclidean", clustering_method = "ward.D2",
           drop_levels = TRUE)
}

df_LQT_corrected_Diff_Oneg_Opos <- rbind(df_LQT_corrected_Up_Oneg_Opos, df_LQT_corrected_Down_Oneg_Opos)
i_adult_old <- which(annotation$Age %in% c("Adult","Old") & annotation$Cell_type %in% c("CD49bneg","HSC_I","CD49bpos"))
i_adult_old_df <- c(1:length(basic), length(basic) + i_adult_old)

df_LQT_corrected_Diff_Oneg_Opos[,i_adult_old_df] %>%
  Row_norm(annotation[i_adult_old,], basic) %>%
  Heat_map_3(annotation[i_adult_old,], basic, "Old subtype comparison including Itokawa",
             colours_Celltypes = colours_Celltypes[c(1,3,2)], colours_Age = colours_Age[2:3])

setEPS()
postscript("../plots/ATACseq_analysis/S9b_Heatmap_Diff_OnegvsOpos_with_Itokawa.eps", width=9, height=6)
df_LQT_corrected_Diff_Oneg_Opos[,i_adult_old_df] %>%
  Row_norm(annotation[i_adult_old,], basic) %>%
  Heat_map_3(annotation[i_adult_old,], basic, "Old subtype comparison including Itokawa",
             colours_Celltypes = colours_Celltypes[c(1,3,2)], colours_Age = colours_Age[2:3])
dev.off()
```